Impact of Demographics and Skills of the Observer on Visual Statistical Inference

### 2013-05-29 sociology_paper.tex
Introduction updated

\begin {abstract}  
Visual Inference is dependent on the careful evaluation of the lineups by individual observers. Each individual is different in their cognitive phycology and judiciousness which can affect the power of visual inference. To estimate the power of visual inference this can be controlled by getting evaluations from multiple people from a diverse population. But the other factor that may affect the power of visual inference is the demographics and skills of the observer. In this paper we examine this in details. The simulation experiments suggest that individual skill as well as demographics are very significant for the power of visual inference. We examine the evidence of learning or getting skilled while sequentially evaluating multiple lineups which enables them evaluate a lineup faster in the later ateempts. The experimental data suggest that there is no learning effect in terms of correct response. The location of actual plot in the lineups does not have any significant effect as well.

@@

The fundamental concept of visual statistical inference is introduced by \citet{buja:2009} and later these concepts are validated by \citet{majumder:2012}.

%\section{Kodu} Tui ekta chor. Tore kilai.

\subsection{Visual Inference} A short introduction of visual inferential procedure.
\subsection{Estimation of Power of Visual Inference} Discussion of the methods for estimating the power of visual inference \citep{majumder:2012}.

\section{Factors Affecting the Power of Visual Inference} A brief description of the factors that may affect the power of visual inference.

\subsection{Choice of Visual Test Statistic} Visual test statistics as defined in \citep{majumder:2012} serve two main purposes. One is to display a prominent pattern that should direct the human observers to select the actual plot in the lineup when the null hypothesis is not true. The another feature is that it should not show any strange pattern when null hypothesis is true, i.e., it should show similar pattern that null plots may have. Thus a visual test statistic is highly associated with the hypothesis. To achieve these purposes it is very important to decide which plot type and plot features should be adopted. As discussed in \citep{majumder:2012} following grammar of graphics developed  by \citet{wilkinson:1999} and later improved by \citet{hadley:2009} can give the task a standard control.

Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots. It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters \citep{majumder:2012} .  \citep{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from another.

@@

\subsection{Question that Human Observer Answers} When a observer is prepared to evaluate a lineup, there needs to be a guideline what exactly should be looked for in a lineup. The researcher knows about the hypothesis but the observer does not know the underlying details of the lineup. Thus the researcher needs to ask a question observer is to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments done by  \citet{majumder:2012}. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

These questions are very crucial for the power of visual test. They help observer think in a controlled direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These question help observer not to be misguided by those patterns. To review further on this \citet{majumder:2012} have also asked why the observers choose a specific plot.

\subsection{Observer Personality} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  notice that some people spend a lot of time to decide no matter whether the lineup is difficult or easy while some simply glance at lineups to make a decision. This influences the response of the observer and we see the subject specific variation in the power of visual test \citep{majumder:2012}.

@@

\subsection{Demographics of Observer}  Age, gender and location of the observer.
\subsection{Skills of Observer} Knowledge about statistical plot as well as education level.

\section{Simulation Experiment for Learning Effect}

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=3in]{rejected_task.pdf}
      \includegraphics[width=3in]{task_duration.pdf} 
   \caption{Percentage of rejected tasks and duration in hour per 100 tasks for each of the 10 experiments. Most of the tasks got rejected for box plot experiment.  Even though the sine illusion experiment took longest to finish the rejection rate is lowest for this experiment.}
   \label{fig:task_duration}
\end{figure}

\subsection{Estimation of sample size}  How the sample size is estimated
\subsection{Data collection methods}  Description of the simulation experiment giving focus on demographics and skills of experimental subjects.
\subsection{Amazon Mechanical Turk}

@@

Because of its convenience it is getting popular for scientific research study. In comparison with a lab study \cite{suri:2010} perform the same study using MTurk and demonstrate that their study results are as good as the lab study results even though MTurk study required less time and cost while provided more convenience. \cite{majumder:2012} recruited people from MTurk for their simulation study in estimating the power of visual statistical inference. They have done numerous pilot studies in lab before doing actual MTurk study and found similar results. \cite{mason:2012} explains various features of MTurk and describes how it can be used as part of human behavioral study.

\section{Results}
\subsection{Overview of the Data}
\subsection{Diversity of the Experimental Subjects} Amazon Mechanical Turk \cite{turk} web site is a good source for recruiting people to evaluation lineups. It is a source of diverse workers who can give us feedback on visual inspection.  Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

\subsection{How people pick the data plot} In the experiment setup we have a question for the turk observers to choose the reason for their selection of a particular data plot. It is shown by \cite{majumder:2012} that these choice reasons reveal the way people picks the data plot in the lineup. To investigate this further we added free text input option for their choice reason instead of some fixed reasons to select from. In this experiment people could write whatever they think their reasons for choice are. Figure \ref{fig:wordle} shows the words used to explain their reasons for choice. The most common words used to explain their choice are points and green which indicates two important feature of a plot. One is the indicator of plot aesthetics and the other is the color. Spread, steepest, line and apart are some other important words used frequently. Spread and apart are indicative of variability in the data. Steepest and line indicates some sort of systematic pattern in the data.

@@

To accomplish this we fit generalized mixed effect model presented in \cite{majumder:2012} with lineups and subjects as random effect and no fixed effect. This gives the estimation of proportion correct responses  and we obtain residuals of the fitted model. The residuals give variations in proportion correct adjusted for lineup difficulties and individual skills. If there is any learning trend over different attempt $A_k$, the residuals should have that information since $A_k$ was not included as a covariate in the model. %At this point we focus on absolute residuals over different attempts to study the deviations due to attempts. An smaller absolute residual at attempt $A_k$ should indicate learning from all other attempts less than $k$.

>>>>

\begin {abstract}  
Visual Statistical Inference is dependent on the careful evaluation of the lineups by individual observers. Each individual is different in their cognitive phycology and judiciousness which can affect the power of visual inference. To estimate the power of visual inference this can be controlled by getting evaluations from multiple people from a diverse population. But the other factor that may affect the power of visual inference is the demographics and skills of the observer. In this paper we examine this in details. The simulation experiments suggest that individual skill as well as demographics are very significant for the power of visual inference. We examine the evidence of learning or getting skilled while sequentially evaluating multiple lineups which enables them evaluate a lineup faster in the later attempts. The experimental data suggest that there is no learning effect in terms of correct response. The location of actual plot in the lineups does not have any significant effect as well.

@@

Statistical graphics can be used to test the significance of findings during the exploratory data analysis. The method is now termed as visual statistical inference which is discussed in the following section.

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as conventional test while in some scenarios, the power exceeds the conventional test power.

The test statistic used in visual inference is a plot of the data. This plot is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the observed plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

A lineup can be evaluated by a single person or multiple observers. Based on the observer feedback the $p$-value is computed \citep{majumder:2013} and finally a decision is made about the hypothesis being tested.

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=6.5in]{plot_turk9_geno_4_3.png} 
   \caption{A lineup of 20 plots. One of the plots is observed data and rest of the 19 plots are obtained by randomly permuting the group structure. Can you identify the observed data?}
   \label{fig:lineup_geno}
\end{figure}

\subsection{Performance of the observer} Observer who evaluates a lineup do not necessarily be aware of the data underlying a lineup. In response to a specific question the observer is to make a choice based on what he or she sees in the lineup. The choice could be right or wrong depending on whether the data plot is selected or not. Under null hypothesis, all the null plots look similar to what the observed data plot appears to be. This would make it harder to detect the observed data plot in the lineup. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup which is 20 in this case, there is a 1/20 chance that the observer would pick the actual plot. This proportion is associated with the type-I error probability. On the other hand if null hypothesis is not true, the observed plot would be different than the null plots making it easier to be distinguished. When multiple observers evaluate a lineup, the proportion of correct response can be used to estimate the power of the visual test \citep{majumder:2013}.

For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observe evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for this before estimating the observer performance.

Some of the factors that may affect the performance of the observer are discussed in Section \ref{sec:factor_performance}. In this paper we focus on estimating the learning effect of the observer in terms of both proportion correct and time taken to evaluate a lineup through the attempts made during the evaluation of multiple plots. We also estimate the location effect of the observed data plot in a lineup.  The other factors are controlled in simulation experiments so that learning effect and location effect can be estimated.

This paper is organized as follows; Section \ref{sec:factor_performance} describes some of the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the design of the simulation experiments and methods to estimate the learning trend of the observer and location effect of the observed plot in a lineup. Section \ref{sec:result_socio} presents the analysis of the experimental data and the results.

\section{Factors Affecting the performance of the observer} \label{sec:factor_performance}

A brief description of the factors that may affect the power of visual inference.

\subsection{Signal in the data}

The power can be estimated from the proportion of correct evaluations. Thus power could be used as a measure of performance of the observer.

\subsection{Choice of Visual Test Statistic} Visual test statistics as defined in \citep{majumder:2013} serve two main purposes. One is to display a prominent pattern that should direct the human observers to select the actual plot in the lineup when the null hypothesis is not true. The another feature is that it should not show any strange pattern when null hypothesis is true, i.e., it should show similar pattern that null plots may have. Thus a visual test statistic is highly associated with the hypothesis. To achieve these purposes it is very important to decide which plot type and plot features should be adopted. As discussed in \citep{majumder:2013} following grammar of graphics developed  by \citet{wilkinson:1999} and later improved by \citet{hadley:2009} can give the task a standard control.

Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots. It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters \citep{majumder:2013} .  \citep{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from another.

@@

\subsection{Question that Human Observer Answers} When a observer is prepared to evaluate a lineup, there needs to be a guideline what exactly should be looked for in a lineup. The researcher knows about the hypothesis but the observer does not know the underlying details of the lineup. Thus the researcher needs to ask a question observer is to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments done by  \citet{majumder:2013}. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

These questions are very crucial for the power of visual test. They help observer think in a controlled direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These question help observer not to be misguided by those patterns. To review further on this \citet{majumder:2013} have also asked why the observers choose a specific plot.

%\subsection{Observer Personality}

@@

\subsection{Demographics of Observer}  Age, gender and location of the observer.

\subsection{Individual Performance of the Observer} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  notice that some people spend a lot of time to decide no matter whether the lineup is difficult or easy while some simply glance at lineups to make a decision. This influences the response of the observer and we see the subject specific variation in the power of visual test \citep{majumder:2013}.

\section{Simulation Experiment and Methods}\label{sec:exp_design}
The performance of the observer while evaluating a lineup would depend on the factors described in Section \ref{sec:facotr_performance}. In this section we present the experimental design to control some  of them so that learning effect of observer and location effect of a plot in the lineup can be estimated.

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=3in]{rejected_task.pdf}
      \includegraphics[width=3in]{task_duration.pdf} 
   \caption{Percentage of rejected tasks and duration of each experiment in hour per 100 tasks for each of the 10 experiments. Most of the tasks got rejected for box plot experiment.  Even though the sine illusion experiment took longest to finish the rejection rate is lowest for this experiment.}
   \label{fig:task_duration}
\end{figure}

%\subsection{Estimation of sample size}  How the sample size is estimated
\subsection{Data collection methods}  Description of the simulation experiment giving focus on demographics and skills of experimental subjects.
\subsection{Amazon Mechanical Turk}

@@

Because of its convenience it is getting popular for scientific research study. In comparison with a lab study \cite{suri:2010} perform the same study using MTurk and demonstrate that their study results are as good as the lab study results even though MTurk study required less time and cost while provided more convenience. \cite{majumder:2013} recruited people from MTurk for their simulation study in estimating the power of visual statistical inference. They have done numerous pilot studies in lab before doing actual MTurk study and found similar results. \cite{mason:2012} explains various features of MTurk and describes how it can be used as part of human behavioral study.

\subsection{Model to Estimate Learning Trend}

\subsection{Model to Estimate Location Effect}

\section{Results}\label{sec:result_socio}
\subsection{Overview of the Data}
\subsection{Diversity of the Experimental Subjects} Amazon Mechanical Turk \cite{turk} web site is a good source for recruiting people to evaluation lineups. It is a source of diverse workers who can give us feedback on visual inspection.  Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

\subsection{How people pick the data plot} In the experiment setup we have a question for the turk observers to choose the reason for their selection of a particular data plot. It is shown by \cite{majumder:2013} that these choice reasons reveal the way people picks the data plot in the lineup. To investigate this further we added free text input option for their choice reason instead of some fixed reasons to select from. In this experiment people could write whatever they think their reasons for choice are. Figure \ref{fig:wordle} shows the words used to explain their reasons for choice. The most common words used to explain their choice are points and green which indicates two important feature of a plot. One is the indicator of plot aesthetics and the other is the color. Spread, steepest, line and apart are some other important words used frequently. Spread and apart are indicative of variability in the data. Steepest and line indicates some sort of systematic pattern in the data.

@@

To accomplish this we fit generalized mixed effect model presented in \cite{majumder:2013} with lineups and subjects as random effect and no fixed effect. This gives the estimation of proportion correct responses  and we obtain residuals of the fitted model. The residuals give variations in proportion correct adjusted for lineup difficulties and individual skills. If there is any learning trend over different attempt $A_k$, the residuals should have that information since $A_k$ was not included as a covariate in the model. %At this point we focus on absolute residuals over different attempts to study the deviations due to attempts. An smaller absolute residual at attempt $A_k$ should indicate learning from all other attempts less than $k$.
